# EPA Ecoregions {#sec-EPAEcoregions}  

For some geological classification that isn't based on watershed boundaries, we turn to the [EPA's Ecoregions](https://www.epa.gov/eco-research/ecoregions).  


Intent here is to list which ecoregions are in our watershed, and which are in each basin (using MDEQ's divisions). And maybe how much are of each basin belongs to each ecoregion. Plus show a map.  

```{r}
library(tidyverse)
library(sf)
library(tmap)
library(msepBoundaries)

# render the maps interactively or statically
out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

if (out_format == "html") {
  message("Setting tmap mode to 'view'")
  tmap_mode("view")
} else {
  message("Setting tmap mode to 'plot'")
  tmap_mode("plot")
}

```

```{r}

fl <- here::here("data", "processed",
                 "EPA_EcoRegion.gpkg")

all3 <- st_read(fl, layer = "level3_lamsal", quiet = TRUE)
msep3 <- st_read(fl, layer = "level3_mssoundwatershed", quiet = TRUE)
msep4 <- st_read(fl, layer = "level4_mssoundwatershed", quiet = TRUE)

msep_boundary <- outline_full |> 
    st_transform(crs = st_crs(all3))
```


## Level 3 Ecoregions  

```{r}
#| label: fig-l3ecoall
#| fig-cap: "Level 3 Ecoregions in Louisiana, Mississippi, and Alabama; the Mississippi Sound Watershed is outlined in black."

tm_shape(all3) +
    tm_basemap("CartoDB.Positron") +
    tm_polygons(fill = "L3_KEY",
             fill_alpha = 0.8,
             fill.scale = tm_scale(values = "viridis"),
             fill.legend = tm_legend(
                 title = "Level 3 Ecoregion Code"
             )) +
    tm_shape(msep_boundary) +
    tm_lines(col = "black",
             lwd = 3)
```

The Mississippi Sound Watershed contains 4 level-3 ecoregions:  

-  65 Southeastern Plains  
-  73 Mississippi Alluvial Plain  (in the very southwest corner)  
-  74 Mississippi Valley Loess Plains  
-  75 Southern Coastal Plain  

```{r}
#| label: fig-l3ecomssnd
#| fig-cap: "Level 3 Ecoregions in the Mississippi Sound Watershed only."

tm_shape(msep3) +
    tm_basemap("CartoDB.Positron") +
    tm_polygons(fill = "L3_KEY",
             fill_alpha = 0.8,
             fill.scale = tm_scale(values = "viridis"),
             fill.legend = tm_legend(
                 title = "Level 3 Ecoregion Code"
             )) +
    tm_shape(msep_boundary) +
    tm_lines(col = "black")
```


## Level 4 Ecoregions  

```{r}
#| label: fig-l4ecomssnd
#| fig-cap: "Level 4 Ecoregions in the Mississippi Sound Watershed."

tm_shape(msep4) +
    tm_basemap("CartoDB.Positron") +
    tm_polygons(fill = "L4_KEY",
             fill_alpha = 0.8,
             fill.scale = tm_scale(values = "viridis"),
             fill.legend = tm_legend(
                 title = "Level 4 Ecoregion Code"
             )) +
    tm_shape(msep_boundary) +
    tm_lines(col = "black") 
```


### About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingEPAEcoregions].



