# Land Use / Land Cover {#sec-lulc}  

The following information was generated using 2023 data from the [USGS Annual National Land Cover Database](https://www.usgs.gov/centers/eros/science/annual-national-land-cover-database). See @sec-processingLULC for more detail.  

```{r}
library(tidyverse)
library(sf)
library(terra)
library(tmap)
library(tidyterra)
library(tigris)
options(tigris_use_cache = TRUE)
```

```{r}
# render the maps interactively or statically
out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

if (out_format == "html") {
  message("Setting tmap mode to 'view'")
  tmap_mode("view")
} else {
  message("Setting tmap mode to 'plot'")
  tmap_mode("plot")
}
```

```{r}
fl <- here::here("data", "processed",
                 "NLCD_MSEP.tif")

lulc <- rast(fl)
lulc_types <- as.data.frame(levels(lulc))
```

```{r}
# state shapes
ms_sf <- states(cb = TRUE, progress_bar = FALSE) |>
    filter(NAME == "Mississippi") |> 
    st_transform(crs = st_crs(lulc))
alla_sf <- states(cb = TRUE, progress_bar = FALSE) |>
    filter(NAME %in% c("Alabama", "Louisiana")) |> 
    st_transform(crs = st_crs(lulc))
```


## Summary of Land Cover Types  

```{r}
type_covers <- table(values(lulc)) |> 
    as.data.frame() 

type_covers <- type_covers |> 
    mutate(value = as.numeric(as.character(Var1))) |>
    select(-Var1) |> 
    left_join(lulc_types)

sumpixels <- sum(type_covers$Freq)

type_covers <- type_covers |> 
    dplyr::mutate(pct_cover = round(100 * Freq / sumpixels, 1))
```

```{r}
type_covers |> 
    select(value,
           `Land Cover Class` = type,
           `Percent of Watershed` = pct_cover) |> 
    gt::gt()
```


## Maps  

```{r}
#| label: fig-lc2023Static
#| fig-cap: "2023 Land Cover designations from the USGS National Land Cover Database."

tmap_mode("plot")

tm <- tm_shape(lulc) +
    tm_raster(title = "Land Cover Class") +
    tm_basemap("CartoDB.Positron") +
    tm_shape(alla_sf) +
    tm_lines("gray10") +
    tm_shape(ms_sf) +
    tm_lines("gray10") +
    tm_layout(legend.outside = TRUE) 

tm
```


```{r}
#| eval: false

# static version of bigger view, to save and use in ccmp

# first shrink it a bit
lulc_view <- terra::aggregate(lulc, fact = 8, fun = modal, na.rm = TRUE)
lulc_view <- as.factor(lulc_view)
coltab(lulc_view) <- coltab(lulc)
levels(lulc_view) <- lulc_types %>%
  select(value, type) %>%
  rename(ID = value)

# current bounding box
current_bbox <- tmaptools::bb(alla_sf)
xrange <- current_bbox[3] - current_bbox[1]
yrange <- current_bbox[4] - current_bbox[2]

# base the new bounding box on the current one because of CRS,
# and just fiddle with the values
newbbox <- current_bbox
# newbbox[1] <- current_bbox[1] + 0.18*xrange   # xmin
# newbbox[2] <- current_bbox[2] + 0.1*yrange    # ymin
# newbbox[3] <- current_bbox[3] - 0.18*xrange   # xmax
# newbbox[4] <- current_bbox[4] - 0.25*yrange   # ymax
# 
# tm_out <- tm_shape(lulc_view, bbox = newbbox) +
#     tm_raster(title = "Land Cover Class") +
#     tm_basemap("CartoDB.Positron") +
#     tm_shape(alla_sf) +
#     tm_lines("gray10") +
#     tm_shape(ms_sf) +
#     tm_lines("gray10")

newbbox[1] <- current_bbox[1] + 0.25*xrange   # xmin
newbbox[2] <- current_bbox[2] + 0.15*yrange    # ymin
newbbox[3] <- current_bbox[3] - 0.25*xrange   # xmax
newbbox[4] <- current_bbox[4] - 0.25*yrange   # ymax

tm_out <- tm_shape(lulc_view, bbox = newbbox) +
    tm_raster(title = "Land Cover") +
    tm_shape(msep) +
    tm_lines("gray") +
    tm_basemap("Esri.WorldTopoMap") +
    tm_shape(alla_sf) +
    tm_lines("gray10") +
    tm_shape(ms_sf) +
    tm_lines("gray10") 

tmap_save(tm_out, here::here("LULC2023.png"))
```



### Interactive  


The pixels in this map have been aggregated so that 1 pixel represents each original 8x8 pixel grid (using the mode; the category that showed up most frequently in the grid). This was done to make it small enough to map interactively.  

```{r}
#| label: fig-lc2023Interactive
#| fig-cap: "2023 Land Cover designations from the USGS National Land Cover Database. Interactive when using the html format of this book."

lulc_view <- terra::aggregate(lulc, fact = 8, fun = modal, na.rm = TRUE)

lulc_view <- as.factor(lulc_view)
coltab(lulc_view) <- coltab(lulc)
levels(lulc_view) <- lulc_types %>%
  select(value, type) %>%
  rename(ID = value)

tm <- tm_shape(lulc_view) +
    tm_basemap("CartoDB.Positron") +
    tm_raster(title = "Land Cover Class") +
    tm_shape(alla_sf) +
    tm_lines("gray10") +
    tm_shape(ms_sf) +
    tm_lines("gray10") +
    tm_layout(legend.outside = TRUE)

if (out_format == "html") {
  message("Setting tmap mode to 'view'")
  tmap_mode("view")
} else {
  message("Setting tmap mode to 'plot'")
  tmap_mode("plot")
}

tm
```




## About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingLULC].