# Population  {#sec-population} 

The values calculated below are based on 2020 US Census data. The population estimate for the entire state of Mississippi was ~2.96 million in 2020.  

```{r}
library(terra)
library(msepBoundaries)
library(sf)
in_file <- here::here("data",
                       "processed",
                       "Population_Dasymetric_2020.tif")
```

```{r}
pop2020 <- rast(in_file)
```

```{r}
# get state and MS Sound boundaries
# and put them into the same CRS as the population file

# entire watershed
msep_bound <- outline_full |> 
    st_transform(crs = st_crs(pop2020))

# watershed only inside MS
msep_ms <- outline_ms_full |> 
    st_transform(crs = st_crs(pop2020))
```

```{r}
# do the cropping/masking
# entire watershed
pop_watershed <- mask(pop2020, msep_bound,
                      progress = FALSE)
# watershed within MS
pop_msep_msonly <- mask(pop2020, msep_ms,
                        progress = FALSE)
```


## Entire watershed  

This estimate includes the portions of the watershed in Louisiana and Alabama:    

<p class = "summary-numberA">
**`r prettyNum(sum(values(pop_watershed), na.rm = TRUE), big.mark = ",")`**
</p>


## Watershed in the state of MS  

This estimate includes the state of Mississippi only:    

<p class = "summary-numberA">
**`r prettyNum(sum(values(pop_msep_msonly), na.rm = TRUE), big.mark = ",")`**
</p>



## About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingPopulation].

