# Precipitation {#sec-Precipitation}  

The maps and information below were generated from gridded (spatial) climate normals: values averaged over 15, 30, or 100 years. Data were obtained from the National Centers for Environmental Information (NCEI).  

```{r}
library(sf)
library(terra)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tidyterra)
library(tigris)
options(tigris_use_cache = TRUE)
library(here)
library(msepBoundaries)

in_path <- here::here("data", "processed")

lamsal <- states(cb = TRUE) |> 
    filter(NAME %in% c("Louisiana",
                       "Mississippi",
                       "Alabama"))
```

```{r}
dat_30yr <- rast(here(in_path, "prcp_30yrNormals.nc"))
monthly_30yr <- dat_30yr["mlyprcp_norm"]
names(monthly_30yr) <- month.abb
annual_30yr <- dat_30yr["annprcp_norm"]
```

```{r}
dat_15yr <- rast(here(in_path, "prcp_15yrNormals.nc"))
monthly_15yr <- dat_15yr["mlyprcp_norm"]
names(monthly_15yr) <- month.abb
annual_15yr <- dat_15yr["annprcp_norm"]
```

```{r}
dat_100yr <- rast(here(in_path, "prcp_100yrBaseline.nc"))
monthly_100yr <- dat_100yr["mlyprcp_norm"]
names(monthly_100yr) <- month.abb
annual_100yr <- dat_100yr["annprcp_norm"]
```

```{r}
annual_combined <- c(annual_100yr, annual_30yr, annual_15yr)
names(annual_combined) <- c("100-year baseline \n(1901-2000)",
                            "30-year normal \n(1991-2020)",
                            "15-year normal \n(2006-2020)")
```

```{r}
lamsal <- st_transform(lamsal, crs = st_crs(annual_combined))
msep_boundary <- st_transform(outline_full, crs = st_crs(annual_combined))
```

```{r}
# stats from 30-year normals
msep_30yr <- crop(annual_30yr, msep_boundary)
msep_30yr <- mask(msep_30yr, msep_boundary)

# get data frames to summarize
msep_30yr <- values(msep_30yr) |> 
    na.omit() |> 
    sort()


n_pixels <- 50
# ms sound watershed
msep_lowest <- head(msep_30yr, n_pixels)
msep_highest <- tail(msep_30yr, n_pixels)

msep_min <- round(mean(msep_lowest))
msep_max <- round(mean(msep_highest))
msep_mean <- round(mean(msep_30yr))
msep_median <- round(median(msep_30yr))
```

```{r}
# 100-yr comps
msep_100yr <- crop(annual_100yr, msep_boundary)
msep_100yr <- mask(msep_100yr, msep_boundary)

msep_100yr <- values(msep_100yr) |> 
    na.omit() |> 
    sort()

# ms sound watershed
msep_lowest_100yr <- head(msep_100yr, n_pixels)
msep_highest_100yr <- tail(msep_100yr, n_pixels)

msep_min_100yr <- round(mean(msep_lowest_100yr))
msep_max_100yr <- round(mean(msep_highest_100yr))
msep_mean_100yr <- round(mean(msep_100yr))
msep_median_100yr <- round(median(msep_100yr))
```


```{r}
# monthly info
msep_monthly <- crop(monthly_30yr, msep_boundary)
msep_monthly <- mask(msep_monthly, msep_boundary)

msep_july <- values(msep_monthly["Jul"]) |> 
    na.omit() |> 
    sort()

# ms sound watershed
msep_highest_july <- tail(msep_july, n_pixels)
msep_max_july <- round(mean(msep_highest_july), 1)
```

```{r}
rm(annual_100yr, annual_30yr, annual_15yr,
   dat_100yr, dat_30yr, dat_15yr)

```

```{r}
#| eval: false

# to help pick limits for color scales
min(values(monthly_30yr), na.rm = TRUE)  # 2.67
max(values(monthly_30yr), na.rm = TRUE)  # 8.21

min(values(annual_combined), na.rm = TRUE)  # 47.56
max(values(annual_combined), na.rm = TRUE)  # 67.63
```


```{r}
breaks_monthly <- seq(2.5, 8.5, by = 1)
colors_monthly <- colorRampPalette(brewer.pal(9, "GnBu"))(length(breaks_monthly) - 1) 

breaks_annual <- seq(45, 70, by = 5)
colors_annual <- colorRampPalette(brewer.pal(9, "GnBu"))(length(breaks_annual) - 1) 
```

```{r}
col_msep <- "#D94F02"

```

## Annual Totals  

The coast receives more rain annually than more inland portions of Louisiana, Mississippi, and Alabama (@fig-prcpAnnual). Based on 30-year climate normals (1991-2020), median annual precipitation in the Mississippi Sound Watershed is `r msep_median` inches. The lowest annual totals, around `r msep_min` inches per year, occur in the northern part of the watershed, with coastal areas receiving around `r msep_max` inches per year. These 30-year normals are somewhat higher than the 100-year baseline period of 1901-2000 (min: `r msep_min_100yr` in; median: `r msep_median_100yr` in; max: `r msep_max_100yr` in).

```{r}
#| label: fig-prcpAnnual
#| fig-cap: "Annual Precipitation averaged over 100, 30, and 15 years."
#| fig-width: 8
#| fig-height: 5

p <- ggplot() +
    geom_spatraster(data = annual_combined) +
    facet_wrap(~lyr,
               ncol = 3) +
    scale_fill_stepsn(
        colours = colors_annual,
        breaks = breaks_annual,
        limits = c(min(breaks_annual), max(breaks_annual)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound Watershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound Watershed" = col_msep)) +
    theme_void() +
    theme(legend.position = "bottom",
          strip.text = element_text(size = rel(0.9), 
                                    face = "bold")) +
    labs(fill = "Inches",
         col = "",
         caption = "Data source: National Centers for Environmental Information")
p
```


## Monthly Totals (30-year normal)  

Seasonal patterns of precipitation also differ throughout the Mississippi Sound Watershed (@fig-prcpMonthly). Coastal areas receive most annual precipitation from June through August (up to `r msep_max_july` inches along the coast in July), while more inland areas receive the most from winter through spring. Tropical storms and hurricanes can and do cause higher precipitation totals than these averages; the peak of hurricane season is typically September-October.   

```{r}
#| label: fig-prcpMonthly
#| fig-cap: "Monthly Precipitation, averaged over 30 years."
#| fig-width: 8
#| fig-height: 6

p <- ggplot() +
    geom_spatraster(data = monthly_30yr) +
    facet_wrap(~lyr) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(legend.position = "right",
          strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = "Inches",
         col = "",
         caption = "Data source: National Centers for Environmental Information")
p
```




## About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingPrecipitation].
