# Precipitation {#sec-Precipitation}  

The maps and information below were generated from gridded (spatial) climate normals: values averaged over 15, 30, or 100 years. Data were obtained from the National Centers for Environmental Information (NCEI).  

```{r}
library(sf)
library(terra)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tidyterra)
library(tigris)
options(tigris_use_cache = TRUE)
library(here)
library(msepBoundaries)

in_path <- here::here("data", "processed")

lamsal <- states(cb = TRUE) |> 
    filter(NAME %in% c("Louisiana",
                       "Mississippi",
                       "Alabama"))
```

```{r}
dat_30yr <- rast(here(in_path, "prcp_30yrNormals.nc"))
monthly_30yr <- dat_30yr["mlyprcp_norm"]
names(monthly_30yr) <- month.abb
annual_30yr <- dat_30yr["annprcp_norm"]
```

```{r}
dat_15yr <- rast(here(in_path, "prcp_15yrNormals.nc"))
monthly_15yr <- dat_15yr["mlyprcp_norm"]
names(monthly_15yr) <- month.abb
annual_15yr <- dat_15yr["annprcp_norm"]
```

```{r}
dat_100yr <- rast(here(in_path, "prcp_100yrBaseline.nc"))
monthly_100yr <- dat_100yr["mlyprcp_norm"]
names(monthly_100yr) <- month.abb
annual_100yr <- dat_100yr["annprcp_norm"]
```

```{r}
annual_combined <- c(annual_100yr, annual_30yr, annual_15yr)
names(annual_combined) <- c("100-year baseline \n(1901-2000)",
                            "30-year normal \n(1991-2020)",
                            "15-year normal \n(2006-2020)")
```

```{r}
rm(annual_100yr, annual_30yr, annual_15yr,
   dat_100yr, dat_30yr, dat_15yr)

```

```{r}
lamsal <- st_transform(lamsal, crs = st_crs(annual_combined))
msep_boundary <- st_transform(outline_full, crs = st_crs(annual_combined))
```

```{r}
#| eval: false

# to help pick limits for color scales
min(values(monthly_30yr), na.rm = TRUE)  # 2.67
max(values(monthly_30yr), na.rm = TRUE)  # 8.21

min(values(annual_combined), na.rm = TRUE)  # 47.56
max(values(annual_combined), na.rm = TRUE)  # 67.63
```


```{r}
breaks_monthly <- seq(2.5, 8.5, by = 1)
colors_monthly <- colorRampPalette(brewer.pal(9, "GnBu"))(length(breaks_monthly) - 1) 

breaks_annual <- seq(45, 70, by = 5)
colors_annual <- colorRampPalette(brewer.pal(9, "GnBu"))(length(breaks_annual) - 1) 
```


## Annual Totals  

```{r}
#| label: fig-prcpAnnual
#| fig-cap: "Annual Precipitation averaged over 100, 30, and 15 years."
#| fig-width: 8
#| fig-height: 5

p <- ggplot() +
    geom_spatraster(data = annual_combined) +
    facet_wrap(~lyr,
               ncol = 3) +
    scale_fill_stepsn(
        colours = colors_annual,
        breaks = breaks_annual,
        limits = c(min(breaks_annual), max(breaks_annual)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound Watershed"),
            linewidth = 0.7) +
    scale_color_manual(values = c("Mississippi Sound Watershed" = "#D94F02")) +
    theme_void() +
    theme(legend.position = "bottom",
          strip.text = element_text(size = rel(0.9), 
                                    face = "bold")) +
    labs(fill = "Inches",
         col = "",
         caption = "Data source: National Centers for Environmental Information")
p
```


## Monthly Totals (30-year normal)  

```{r}
#| label: fig-prcpMonthly
#| fig-cap: "Monthly Precipitation, averaged over 30 years."
#| fig-width: 8
#| fig-height: 6

p <- ggplot() +
    geom_spatraster(data = monthly_30yr) +
    facet_wrap(~lyr) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.7) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = "#D94F02")) +
    theme_void() +
    theme(legend.position = "right",
          strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = "Inches",
         col = "",
         caption = "Data source: National Centers for Environmental Information")
p
```




## About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingPrecipitation].
