# Salinity {#sec-salinity}  

Data for this section was obtained from the [Gulf Data Atlas](https://gulfatlas.noaa.gov/). This was part of a larger effort to characterize seasonality in Gulf Coast estuaries, by assigning salinity zones in different time periods, based on long-term data. See [Appendix @sec-processingSalinity] for more detail.  

To view more recent data from USGS stations in the Mississippi Sound, check out our interactive [Mississippi Sound Salinity Explorer App](https://cmep-ms.shinyapps.io/SalinityExplorer/)!

```{r}
library(sf)
library(dplyr)
library(msepBoundaries)
library(tmap)

# render the maps interactively or statically
out_format <- knitr::opts_knit$get("rmarkdown.pandoc.to")

if (out_format == "html") {
  message("Setting tmap mode to 'view'")
  tmap_mode("view")
} else {
  message("Setting tmap mode to 'plot'")
  tmap_mode("plot")
}
```

```{r}
tmap_options(component.autoscale = FALSE)
coastal_bb <- st_bbox(c(xmin = -89.80829, xmax = -87.71594,
                   ymin = 29.81824, ymax = 30.89847),
                 crs = st_crs(4269))

msep_outline <- outline_full |> 
    st_transform(crs = st_crs(4269))

# make_sal_map <- function(data, fill_var, map_title) {
#     m <- tm_shape(data, bbox = coastal_bb) +
#         tm_basemap("CartoDB.Positron") +
#         tm_fill(fill = fill_var,
#                 fill.scale = tm_scale_categorical(values = "brewer.yl_gn_bu"),
#                 fill.legend = tm_legend(
#                     title = "Salinity (psu)",
#                     position = tm_pos_in("right", "bottom"),
#                     bg.color = adjustcolor("white", alpha.f = 0.8)
#                 )) +
#         tm_shape(msep_outline) +
#         tm_borders(col = "black", lwd = 2) +
#         tm_title(map_title)
#     
#     m
# }
```

```{r}
dat <- st_read(here::here("data",
                          "processed",
                          "salinity_gulfDataAtlas",
                          "MS_SalSeasons.shp"),
               quiet = TRUE)

# Ensure SAL_LOW and SAL_HIGH are factors with the correct levels
dat$SAL_LOW <- factor(dat$SAL_LOW, levels = c("0-0.5", "0.5-5", "05-15", "15-25", ">25"))
dat$SAL_HIGH <- factor(dat$SAL_HIGH, levels = c("0-0.5", "0.5-5", "05-15", "15-25", ">25"))
```

```{r}
#| label: fig-mapLow
#| fig-cap: "Average salinity (psu) during the low salinity period, February - April. Data layer courtesy of the Gulf Data Atlas."
#| fig-width: 8
#| fig-height: 6.5

# tm_shape(dat, bbox = coastal_bb) +
#         tm_basemap("CartoDB.Positron") +
#         tm_fill(fill = "SAL_LOW",
#                 fill.scale = tm_scale_categorical(values = "brewer.yl_gn_bu"),
#                 fill.legend = tm_legend(
#                     title = "Salinity (psu)",
#                     position = tm_pos_in("right", "bottom"),
#                     bg.color = adjustcolor("white", alpha.f = 0.8)
#                 )) +
#         tm_shape(msep_outline) +
#         tm_borders(col = "black", lwd = 2) +
#         tm_title("Low Salinity Period\n(Feb-Apr)")

tm_shape(dat, bbox = coastal_bb) +
    tm_fill(fill = "SAL_LOW",
            fill.scale = tm_scale_categorical(values = "brewer.yl_gn_bu"),
            fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                # bg.color = adjustcolor("white", alpha.f = 0.8),
                title = "Salinity (psu)")
            ) +
    tm_basemap("CartoDB.Positron") +
    tm_shape(msep_outline) +
    tm_borders(col = "black", lwd = 2)
```

```{r}
#| label: fig-mapHigh
#| fig-cap: "Average salinity (psu) during the high salinity period, August - October. Data layer courtesy of the Gulf Data Atlas."
#| fig-width: 8
#| fig-height: 6.5

# tm_shape(dat, bbox = coastal_bb) +
#         tm_basemap("CartoDB.Positron") +
#         tm_fill(fill = "SAL_HIGH",
#                 fill.scale = tm_scale_categorical(values = "brewer.yl_gn_bu"),
                # fill.legend = tm_legend(
                #     title = "Salinity (psu)",
                #     position = tm_pos_in("right", "bottom"),
                #     bg.color = adjustcolor("white", alpha.f = 0.8)
                # )) +
#         tm_shape(msep_outline) +
#         tm_borders(col = "black", lwd = 2) +
#         tm_title("High Salinity Period\n(Aug-Oct)")

tm_shape(dat, bbox = coastal_bb) +
    tm_fill(fill = "SAL_HIGH",
            fill.scale = tm_scale_categorical(values = "brewer.yl_gn_bu"),
            fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                # bg.color = adjustcolor("white", alpha.f = 0.8),
                title = "Salinity (psu)")
            ) +
    tm_basemap("CartoDB.Positron") +
    tm_shape(msep_outline) +
    tm_borders(col = "black", lwd = 2)

```



## About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingSalinity].