# Temperature {#sec-Temperature} 

The maps and information below were generated from gridded (spatial) climate normals: values averaged over 15, 30, or 100 years. Data were obtained from the National Centers for Environmental Information (NCEI).  

```{r}
library(sf)
library(terra)
library(RColorBrewer)
library(ggplot2)
library(dplyr)
library(tidyterra)
library(tigris)
options(tigris_use_cache = TRUE)
library(here)
library(msepBoundaries)
library(stringr)
library(patchwork)

in_path <- here::here("data", "processed")

lamsal <- states(cb = TRUE) |> 
    filter(NAME %in% c("Louisiana",
                       "Mississippi",
                       "Alabama"))
```

```{r}
dat_30yr <- rast(here(in_path, "tempAvg_30yrNormals.nc"))
monthly_30yr <- dat_30yr["mlytavg_norm"]
names(monthly_30yr) <- month.abb
annual_30yr <- dat_30yr["anntavg_norm"]
```

```{r}
dat_15yr <- rast(here(in_path, "tempAvg_15yrNormals.nc"))
monthly_15yr <- dat_15yr["mlytavg_norm"]
names(monthly_15yr) <- month.abb
annual_15yr <- dat_15yr["anntavg_norm"]
```

```{r}
dat_100yr <- rast(here(in_path, "tempAvg_100yrBaseline.nc"))
monthly_100yr <- dat_100yr["mlytavg_norm"]
names(monthly_100yr) <- month.abb
annual_100yr <- dat_100yr["anntavg_norm"]
```

```{r}
annual_combined <- c(annual_100yr, annual_30yr, annual_15yr)
names(annual_combined) <- c("100-year baseline \n(1901-2000)",
                            "30-year normal \n(1991-2020)",
                            "15-year normal \n(2006-2020)")
```

```{r}
rm(annual_100yr, annual_30yr, annual_15yr,
   dat_100yr, dat_30yr, dat_15yr)

```

```{r}
lamsal <- st_transform(lamsal, crs = st_crs(annual_combined))
msep_boundary <- st_transform(outline_full, crs = st_crs(annual_combined))
```

```{r}
#| eval: false

# to help pick limits for color scales
min(values(monthly_30yr), na.rm = TRUE)  # 38.01
max(values(monthly_30yr), na.rm = TRUE)  # 83.70

min(values(annual_combined), na.rm = TRUE)  # 57.34
max(values(annual_combined), na.rm = TRUE)  # 71.33
```


```{r}
breaks_monthly <- seq(35, 85, by = 5)
labels_monthly <- as.character(breaks_monthly)
labels_monthly[which(str_ends(labels_monthly, "0"))] <- ""
colors_monthly <- colorRampPalette(brewer.pal(9, "YlOrBr"))(length(breaks_monthly) - 1) 

breaks_annual <- seq(55, 73, by = 3)
colors_annual <- colorRampPalette(brewer.pal(9, "YlOrBr"))(length(breaks_annual) - 1) 
```


## Annual Mean Temperature  

```{r}
#| label: fig-tempAnnual
#| fig-cap: "Annual Mean Temperature averaged over 100, 30, and 15 years."
#| fig-width: 8
#| fig-height: 5

p <- ggplot() +
    geom_spatraster(data = annual_combined) +
    facet_wrap(~lyr,
               ncol = 3) +
    scale_fill_stepsn(
        colours = colors_annual,
        breaks = breaks_annual,
        limits = c(min(breaks_annual), max(breaks_annual)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound Watershed"),
            linewidth = 0.7) +
    scale_color_manual(values = c("Mississippi Sound Watershed" = "navy")) +
    theme_void() +
    theme(legend.position = "bottom",
          strip.text = element_text(size = rel(0.9), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         caption = "Data source: National Centers for Environmental Information")
p
```


## Monthly Mean Temperature   

```{r}
col_msep <- "#008B8B"

# steelblue
# this teal works pretty well #008080
# this dark cyan works even better #008B8B
# cadet blue is pretty good too #5F9EA0
```


### Cool Months  

```{r}
p30 <- ggplot() +
    geom_spatraster(data = monthly_30yr[[c("Nov", "Dec", "Jan", "Feb", "Mar")]]) +
    facet_wrap(~lyr,
               nrow = 1) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        labels = labels_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         subtitle = "30-year normal")
```

```{r}
p15 <- ggplot() +
    geom_spatraster(data = monthly_15yr[[c("Nov", "Dec", "Jan", "Feb", "Mar")]]) +
    facet_wrap(~lyr,
               nrow = 1) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        labels = labels_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         subtitle = "15-year normal")
```


```{r}
p100 <- ggplot() +
    geom_spatraster(data = monthly_100yr[[c("Nov", "Dec", "Jan", "Feb", "Mar")]]) +
    facet_wrap(~lyr,
               nrow = 1) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        labels = labels_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         subtitle = "100-year baseline")

```

```{r}
#| label: fig-tmpNormalsWinter
#| fig-cap: "Comparison of long-term winter temperature averages across a 100-year baseline, and 30- and 15-year normals."


p100 / p30 / p15 +
    plot_layout(guides = "collect") +
    plot_annotation(caption = "Data source: National Centers for Environmental Information") &
    theme(legend.position = "right")
```


### Warm Months  

```{r}
p30 <- ggplot() +
    geom_spatraster(data = monthly_30yr[[c("Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")]]) +
    facet_wrap(~lyr,
               nrow = 1) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        labels = labels_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         subtitle = "30-year normal")
```

```{r}
p15 <- ggplot() +
    geom_spatraster(data = monthly_15yr[[c("Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")]]) +
    facet_wrap(~lyr,
               nrow = 1) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        labels = labels_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         subtitle = "15-year normal")
```


```{r}
p100 <- ggplot() +
    geom_spatraster(data = monthly_100yr[[c("Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")]]) +
    facet_wrap(~lyr,
               nrow = 1) +
    scale_fill_stepsn(
        colours = colors_monthly,
        breaks = breaks_monthly,
        labels = labels_monthly,
        limits = c(min(breaks_monthly), max(breaks_monthly)),
        na.value = NA
    ) +
    geom_sf(data = lamsal,
            fill = NA,
            col = "black") +
    geom_sf(data = msep_boundary,
            fill = NA,
            aes(col = "Mississippi Sound \nWatershed"),
            linewidth = 0.8) +
    scale_color_manual(values = c("Mississippi Sound \nWatershed" = col_msep)) +
    theme_void() +
    theme(strip.text = element_text(size = rel(0.8), 
                                    face = "bold")) +
    labs(fill = expression(degree*F),
         col = "",
         subtitle = "100-year baseline")

```

```{r}
#| label: fig-tmpNormalsSpringSummer
#| fig-cap: "Comparison of long-term spring and summer temperature averages across a 100-year baseline, and 30- and 15-year normals."
#| fig-height: 6
#| fig-width: 8

p100 / p30 / p15 +
    plot_layout(guides = "collect") +
    plot_annotation(caption = "Data source: National Centers for Environmental Information") &
    theme(legend.position = "bottom")
```

## About {.unnumbered}   

**The `.qmd` file** that generated this section was: **`r xfun::with_ext(knitr::current_input(), "qmd")`**. 

For information on data processing, see [Appendix @sec-processingTemperature].
